Migration of Gas Giant Planets in Gravitationally Unstable Disks

### 2011-06-07 Text/migration.tex
Final submitted version of the migration letter.
	modified:   Text/Figures/amt.eps
	new file:   Text/Figures/amt_bw.eps
	modified:   Text/Figures/planeta.eps
	deleted:    Text/Figures/planeta.pdf
	new file:   Text/Figures/planeta_bw.eps
	modified:   Text/Figures/planete.eps
	deleted:    Text/Figures/planete.pdf
	new file:   Text/Figures/planete_bw.eps
	modified:   Text/general.bib
	modified:   Text/migration.tex
	copied:     Text/migration.tex -> Text/migration_apjlsubmission.tex

git-svn-id: http://jaga.astro.indiana.edu/paperrepo/ApJL/Migration2010@86 3b7f7548-5c66-0410-83ab-3997167635fc

\begin{abstract}
Characterization of migration in gravitationally unstable disks is necessary to understand the fate of protoplanets formed by disk instability. As part of a larger study, we are using a 3D radiative hydrodynamics code to investigate how an embedded gas giant planet interacts with a gas disk that undergoes gravitational instabilities (GIs). This Letter presents preliminary results from simulations with a Jupiter-mass planet placed in orbit at 25 AU within a 0.14 $M_{\odot}$ disk. The disk spans 5 to 40 AU around a 1 $M_{\odot}$ star and is initially marginally unstable. In one simulation, the planet is inserted prior to the eruption of GIs; in another, it is inserted only after the disk has settled into a quasi-steady GI-active state, where heating by GIs roughly balances radiative cooling. When the planet is present from the beginning, its own wake stimulates growth of a particular global mode with which it strongly interacts, and the planet plunges inward six AU in about 10$^3$ years. In both cases with embedded planets, there are times when the planet's radial motion is slow and varies in direction. At other times, when the planet appears to be interacting with strong spiral modes, migration both inward and outward can be relatively rapid, covering several AUs over hundreds of years. Migration in both cases appears to stall near the inner Lindblad resonance of a dominant low-order mode. Planet orbit eccentricities fluctuate rapidly between about 0.02 to 0.1 throughout the GI-active phases of the simulations. 
\end{abstract}

@@

Migration of gas giant planets due to interactions with a circumstellar gas disk can play a major role in defining the architecture of planetary systems. Work on migration \citep[see review by][]{papaloizou2007} has included gravitational interaction of planets with both laminar and turbulent disks. However, radiative transport, detailed equations of state (EOS), and the self-gravity of the gas disk have usually been ignored; the effects of a non-isothermal EOS have only recently been included \citep[e.g.,][]{paardekooper2006,paardekooper2010,paardekooper2011}. 
\citet{boss2005} and \citet{mayer2004} examined radial migration of planet-mass fragments in gravitationally unstable disks, but their disks were violently disrupted by fragmentation under conditions (radii $<$ 40 AU, disk-to-star mass ratios $M_d/M_s \sim 0.1$, and stellar mass $M_s \sim 1\;M_{\odot}$) where fragmentation may not actually occur  \citep{rafikov2005,rafikov2007,boley2006,boley2007b,boley2008,forgan2009,cai2010}. More recently, fragmentation into clumps with gas giant or brown dwarf masses has been documented in numerical simulations of disks that are relatively massive ($M_d/M_s \sim$ a few tenths) and spatially extended (outer radii $>$ 50 AU) \citep{krumholz2007,stamatellos2007,stamatellos2009,boley2009,boley2010}, where fragmentation is expected from semi-analytic arguments \citep[e.g.,][]{clarke2009,rafikov2009,dodson2009}. The fate of the clumps then depends in part on their radial migration, which is a chaotic and messy affair in a fragmenting disk \citep[e.g.,][]{boley2009,boley2010,vorobyov2010,boley2010b}. The occurrence of gravitational instabilities (GIs) may be episodic \citep[e.g.,][]{vorobyov2006,vorobyov2010b,zhu2010}. Clumps that survive and contract to the dimensions of young planets can later find themselves in a disk that erupts again into GI activity. As the star/disk system evolves, such a protoplanet may end up in a region of a GI-active disk where fragmentation does not occur.

To improve our understanding of how planets migrate in GI-active disks that are not fragmenting, we have begun a systematic study, using numerical 3D radiative hydrodynamics, where we investigate the effects on both the disk and the planet of inserting a planet-mass object into disks susceptible to GIs. Using techniques deployed in earlier research by our group \citep{pickett2003, mejia2005,cai2006,cai2008,boley2006,boley2007b,michael2010}, we can identify the dominant spiral waves in a simulation and analyze how the waves interact with the planet's motion. Our goal is to determine both the effect of giant planets on GIs and the effect of GIs on planet migration. Because GIs are sensitive to radiative physics, we use a well-tested radiative scheme \citep{boley2007b} and realistic opacities \citep{dalessio2001}.

Section 2 below presents our numerical methods and initial conditions.  We describe the simulation results in Section 3, and discuss them in Section 4.

@@

The model disk, based on an equlibrium model in \citet{pickett2003}, orbits a 1 $M_{\odot}$ star and has a mass $M_d = 0.14 M_{\odot}$, inner and outer radii at 5 and 40 AU, and an initial surface density distribution $\Sigma \sim \varpi^{-1/2}$. The time unit of one ORP (= outer rotation period) is defined as the rotation period of the initial disk at $\varpi \approx$ 32 AU, or about 180 yr. The disk is initially located between radial grid zones 30 and 240 and is close to isentropic, which results in a Toomre-$Q$ distribution with a marginally unstable \citep[see ][]{durisen2007} minimum value of 1.38 at radial grid zone 161 (26.7 AU). The computational grid extends radially to 512 zones to accommodate expansion of the outer disk once GIs become nonlinear. An outflow boundary condition is enforced at the upper vertical grid boundary, the outer radial grid boundary, and an inner radial boundary at 2 AU. To seed nonaxisymmetry, the density is given an initial 0.01 \% random cell-to-cell perturbation.

This Letter presents three simulations. The first, which we call the {\sl fiducial run}, goes from time $t = 0$ to 21 ORP without a planet. In the second simulation, which we call the {\sl t = 0 planet run}, a one Jupiter-mass ($M_J$) planet with a roughly circular midplane orbit is included in the initial $t = 0$ ORP equilibrium disk at 25 AU. In the third simulation, which we call the {\sl t = 10 planet run}, a 1 $M_J$ planet with a similar orbit is inserted into the fiducial run at $t = 10.5$ ORP. These three simulations are part of a larger suite, to be described elsewhere, in which the planet mass is varied. We estimate crude semi-major axes $a$ and eccentricities $e$ from the planet's $\varpi(t)$ by using the maximum and minimum radii, $\varpi_{\rm max}$ and $\varpi_{\rm min}$, attained by the planet over each interval that the planet's $\phi$-coordinate changes by 2$\pi$ (one ``orbit'').  Then, for that orbit, $a = (\varpi_{\rm max}+\varpi_{\rm min})/2$ and $e = (\varpi_{\rm max}-\varpi_{\rm min})/(\varpi_{\rm max}+\varpi_{\rm min})$. These parameters are only indicative; there are strong orbit to orbit variations in $a$ and $e$.

@@

{\it The $t = 10$ Planet Run.} The motion of the planet in this case is more difficult to interpret. We see intervals of fairly rapid inward or outward migration over several orbits, as well as times when radial migration appears to stall. Between $t = 15$ and 19 ORP, the pattern of outward migration for 2 ORP followed by an inward plunge over the next 2 ORP resembles the behavior in the $t = 0$ planet run during the burst, but, in this case, there is no distinct transition between phases of GI activity. Animations of the simulation (available at http://hdl.handle.net/2022/13304) show there is a complex interaction between the planet, its spiral wake, and the global spiral arms caused by the instabilities. Analysis of images and periodicities suggests that, between 15 and 19 ORP, the planet may be in a mean motion resonance with $m = 2$ and 3 patterns, both of which have CR at 27 AU. Over this time interval, the average orbit period of the planet is $\sim$ 0.50 ORP, while both the $m = 2$ and 3 spirals have a pattern period of $\sim 0.74$ ORP, suggesting a 3:2 resonance. Both of these modes are present, at nearly the same periodicities, in the fiducial run. So, in contrast with the $t = 0$ planet run, this planet's rapid phases of radial migration are caused by interaction with strong GI modes that exist independent of the planet.

@@

{\it Summary.} The simulations in this Letter show that a planet of even only 1$M_J$ placed near the $Q$-minimum in a 0.14$M_{\odot}$ disk can hasten the onset of gravitational instabilities and change the nature of the dominant modes during the initial burst. Although the magnitude of radial migration (a few$\times10^{-3}$ AU yr$^{-1}$) is not too different from that computed using the \citet{tanaka2002}  isothermal disk formula based on the Lindblad torques, it can fluctuate in sign on time scales on the order of the orbit period and can be strongly affected by gravitational interactions of the planet with discrete GI-driven spiral modes, leading to phases of rapid radial motion. Planet orbit eccentricities fluctuate between about 0.02 and 0.10. The simulations also suggest that planet migration may stall for long periods near the ILRs of dominant global GI-modes.

@@

The authors would like to thank F.C. Adams and J.C.B. Papaloizou for useful encouragement in this project. The research was conducted with the support of NASA grants from the Origins of Solar Systems  Program (NNG05GN11G and NNX08AK36G). S.M. was supported during portions of this research by an NESSP Graduate Fellowship and by Indiana Space Grant Consortium Graduate Fellowships; A.C.B. is currently a Sagan Postdoctoral Fellow at U. Florida. This material is based upon work supported by the National Science Foundation under Grants No. ACI-0338618l, OCI-0451237, OCI-0535258, OCI-0504075, and CNS-0521433, and through TeraGrid resources provided by Indiana University, PSC, SDSC, and NCSA. TeraGrid systems are hosted by Indiana University, LONI, NCAR, NCSA, NICS, ORNL, PSC, Purdue University, SDSC, TACC, and UC/ANL.. Any opinions, findings and conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the views of the National Science Foundation.

@@

\begin{figure}[t]
\center
\includegraphics[width=11cm]{Figures/amt.eps}
\caption{Global amplitudes of nonaxisymmetric density perturbations $A_m$ as a function of time for individual cos($m\phi$) perturbations. The formula for computing $A_m$ is equation (15) of \citep{boley2006}. Here the integrals are done only over the disk outside 15 AU to suppress contributions from edge effects in the inner disk. The set of red and black curves correspond to the fiducial run (no planet) and the $t = 0$ planet run, respectively.}
\label{fig:Am}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/planeta.eps}
\caption{Plot of the planet's radial position as a function of time $\varpi_p(t)$ for the two planet simulations. The red and black indicate the curves ($\varpi(t)$), points ($a$), and horizontal scales ($t$) that correspond to the $t = 0$ and $t = 10$ planet runs, respectively. The diamonds are the approximate semi-major axes $a$ computed for each 0 to 2$\phi$ change in azimuth of the planet as explained in the text.}
\label{fig:a}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/planete.eps}
\caption{Plot of the eccentricities $e$, computed for each 0 to 2$\phi$ change in azimuth of the planet as explained in the text, as a function of time. The red and black indicate the points ($e$) and horizontal scales ($t$) that correspond to the $t = 0$ and $t = 10$ planet runs, respectively.}
\label{fig:e}
\end{figure}

>>>>

\begin{abstract}
Characterization of migration in gravitationally unstable disks is necessary to understand the fate of protoplanets formed by disk instability. As part of a larger study, we are using a 3D radiative hydrodynamics code to investigate how an embedded gas giant planet interacts with a gas disk that undergoes gravitational instabilities (GIs). This Letter presents results from simulations with a Jupiter-mass planet placed in orbit at 25 AU within a 0.14 $M_{\odot}$ disk. The disk spans 5 to 40 AU around a 1 $M_{\odot}$ star and is initially marginally unstable. In one simulation, the planet is inserted prior to the eruption of GIs; in another, it is inserted only after the disk has settled into a quasi-steady GI-active state, where heating by GIs roughly balances radiative cooling. When the planet is present from the beginning, its own wake stimulates growth of a particular global mode with which it strongly interacts, and the planet plunges inward six AU in about 10$^3$ years. In both cases with embedded planets, there are times when the planet's radial motion is slow and varies in direction. At other times, when the planet appears to be interacting with strong spiral modes, migration both inward and outward can be relatively rapid, covering several AUs over hundreds of years. Migration in both cases appears to stall near the inner Lindblad resonance of a dominant low-order mode. Planet orbit eccentricities fluctuate rapidly between about 0.02 to 0.1 throughout the GI-active phases of the simulations. 
\end{abstract}

@@

Migration of gas giant planets due to interactions with a circumstellar gas disk can play a major role in defining the architecture of planetary systems. Work on migration \citep[see review by][]{papaloizou2007} has included gravitational interaction of planets with both laminar and turbulent disks. However, radiative transport, detailed equations of state (EOS), and the self-gravity of the gas disk have usually been ignored; the effects of a non-isothermal EOS have only recently been included \citep[e.g.,][]{paardekooper2006,paardekooper2010,paardekooper2011}. Emerging studies, such as \citet{baruteau2011} are just beginning to address some of these issues. 
\citet{boss2005} and \citet{mayer2004} examined radial migration of planet-mass fragments in gravitationally unstable disks, but their disks were violently disrupted by fragmentation under conditions (radii $<$ 40 AU, disk-to-star mass ratios $M_d/M_s \sim 0.1$, and stellar mass $M_s \sim 1\;M_{\odot}$) where fragmentation may not actually occur  \citep{rafikov2005,rafikov2007,boley2006,boley2007b,boley2008,forgan2009,cai2010}. More recently, fragmentation into clumps with gas giant or brown dwarf masses has been documented in numerical simulations of disks that are relatively massive ($M_d/M_s \sim$ a few tenths) and spatially extended (outer radii $>$ 50 AU) \citep{krumholz2007,stamatellos2007,stamatellos2009,boley2009,boley2010}, where fragmentation is expected from semi-analytic arguments \citep[e.g.,][]{clarke2009,rafikov2009,dodson2009}. The fate of the clumps then depends in part on their radial migration, which is a chaotic and messy affair in a fragmenting disk \citep[e.g.,][]{boley2009,boley2010,vorobyov2010,boley2010b}. The occurrence of gravitational instabilities (GIs) may be episodic \citep[e.g.,][]{vorobyov2006,vorobyov2010b,zhu2010}. Clumps that survive and contract to the dimensions of young planets can later find themselves in a disk that erupts again into GI activity. As the star/disk system evolves, such a protoplanet may end up in a region of a GI-active disk where fragmentation does not occur.

To improve our understanding of how planets migrate in GI-active disks that are not fragmenting, we have begun a systematic study, using numerical 3D radiative hydrodynamics, where we investigate the effects on both the disk and the planet of inserting a planet-mass object into disks susceptible to GIs. Using techniques developed in earlier research by our group \citep{pickett2003, mejia2005,cai2006,cai2008,boley2006,boley2007b,michael2010}, we can identify the dominant spiral waves in a simulation and analyze how the waves interact with the planet's motion. Our goal is to determine both the effect of giant planets on GIs and the effect of GIs on planet migration. Because GIs are sensitive to radiative physics, we use a well-tested radiative scheme \citep{boley2007b} and realistic opacities \citep{dalessio2001}.

Section 2 below presents our numerical methods and initial conditions.  We describe the simulation results in \S3, and discuss them in \S4.

@@

The model disk, based on an equlibrium model from \citet{pickett2003}, orbits a 1 $M_{\odot}$ star and has a mass $M_d = 0.14 M_{\odot}$, inner and outer radii at 5 and 40 AU, and an initial surface density distribution $\Sigma \sim \varpi^{-1/2}$. The time unit of one ORP (= outer rotation period) is defined as the rotation period of the initial disk at $\varpi \approx$ 32 AU, or about 180 yr. The disk is initially located between radial grid zones 30 and 240 and is close to isentropic, which results in a Toomre-$Q$ distribution with a marginally unstable \citep[see ][]{durisen2007} minimum value of 1.38 at radial grid zone 161 (26.7 AU). The computational grid extends radially to 512 zones to accommodate expansion of the outer disk once GIs become nonlinear. An outflow boundary condition is enforced at the upper vertical grid boundary, the outer radial grid boundary, and an inner radial boundary at 2 AU. To seed nonaxisymmetry, the density distribution is given an initial 0.01 \% random cell-to-cell perturbation.

This Letter presents three simulations. The first, which we call the {\sl fiducial run},  is simulated from time $t = 0$ to 21 ORP without a planet. In the second simulation, which we call the {\sl t = 0 planet run}, a one Jupiter-mass ($M_J$) planet with a roughly circular midplane orbit is included in the initial $t = 0$ ORP equilibrium disk at 25 AU. In the third simulation, which we call the {\sl t = 10 planet run}, a 1 $M_J$ planet with a similar orbit is inserted into the fiducial run at $t = 10.5$ ORP. These three simulations are part of a larger suite, to be described elsewhere, in which the planet mass is varied. We estimate semi-major axes $a$ and eccentricities $e$ from the planet's $\varpi(t)$ by using the maximum and minimum radii, $\varpi_{\rm max}$ and $\varpi_{\rm min}$, over each complete orbit of the planet, where one orbit is $\Delta\phi=2\pi$. Specifically, $a = (\varpi_{\rm max}+\varpi_{\rm min})/2$ and $e = (\varpi_{\rm max}-\varpi_{\rm min})/(\varpi_{\rm max}+\varpi_{\rm min})$. These parameters are intended to highlight the strong orbit-to-orbit variations in the motion of the planet.  A more refined estimate of the orbital elements is unnecessary because the potential of the disk, even if were axisymmetric, will cause departures from Keplerian dynamics.

@@

{\it The $t = 10$ Planet Run.} The motion of the planet in this case is more difficult to interpret. We see intervals of fairly rapid inward or outward migration over several orbits, as well as times when radial migration appears to stall. Between $t = 15$ and 19 ORP, the pattern of outward migration for 2 ORP followed by an inward plunge over the next 2 ORP resembles the behavior in the $t = 0$ planet run during the burst, but, in this case, there is no distinct transition between phases of GI activity. Animations of the evolution of the midplane density (available at http://hdl.handle.net/2022/13304) show there is a complex interaction between the planet, its spiral wake, and the global spiral arms caused by the instabilities. Analysis of images and periodicities suggests that, between 15 and 19 ORP, the planet may be in a mean motion resonance with $m = 2$ and 3 patterns, both of which have CR at 27 AU. Over this time interval, the average orbit period of the planet is $\sim$ 0.50 ORP, while both the $m = 2$ and 3 spirals have a pattern period of $\sim 0.74$ ORP, suggesting a 3:2 resonance. Both of these modes are present, at nearly the same periodicities, in the fiducial run. So, in contrast with the $t = 0$ planet run, this planet's rapid phases of radial migration are caused by interaction with strong GI modes that exist independent of the planet.

@@

{\it Summary.} The simulations in this Letter show that a planet of only 1$M_J$ placed near the $Q$-minimum in a 0.14$M_{\odot}$ disk can hasten the onset of gravitational instabilities and change the nature of the dominant modes during the initial burst. Although the magnitude of radial migration (a few$\times10^{-3}$ AU yr$^{-1}$) is not too different from that computed using the \citet{tanaka2002}  isothermal disk formula based on the Lindblad torques, it can fluctuate in sign on time scales on the order of the orbit period and can be strongly affected by gravitational interactions of the planet with discrete GI-driven spiral modes, leading to phases of rapid radial motion. Planet orbit eccentricities fluctuate between about 0.02 and 0.10. The simulations also suggest that planet migration may stall for long periods near the ILRs of dominant global GI-modes.

@@

The authors would like to thank F.C. Adams and J.C.B. Papaloizou for useful encouragement in this project. The research was conducted with the support of NASA grants from the Origins of Solar Systems  Program (NNG05GN11G and NNX08AK36G). S.M. was supported during portions of this research by an NESSP Graduate Fellowship and by Indiana Space Grant Consortium Graduate Fellowships; A.C.B. is currently a Sagan Postdoctoral Fellow at U. Florida. This material is based upon work supported by the National Science Foundation under Grants No. ACI-0338618l, OCI-0451237, OCI-0535258, OCI-0504075, and CNS-0521433, and through TeraGrid resources provided by Indiana University, PSC, SDSC, and NCSA. TeraGrid systems are hosted by Indiana University, LONI, NCAR, NCSA, NICS, ORNL, PSC, Purdue University, SDSC, TACC, and UC/ANL. Any opinions, findings and conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the views of the National Science Foundation.

@@

\begin{figure}[t]
\center
\includegraphics[width=11cm]{Figures/amt.eps}
\caption{(Color figure available in the online version) Global amplitudes of nonaxisymmetric density perturbations $A_m$ as a function of time for individual cos($m\phi$) perturbations. The formula for computing $A_m$ is equation (15) of \citep{boley2006}. Here the integrals are done only over the disk outside 15 AU to suppress contributions from edge effects in the inner disk. The top and bottom panels correspond to the fiducial run (no planet) and the $t = 0$ planet run, respectively.}
\label{fig:Am}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/planeta.eps}
\caption{Plot of the planet's radial position as a function of time $\varpi_p(t)$ for the two planet simulations. The dashed-diamond and solid-star lines (colored red and black in the online version) indicate the curves ($\varpi(t)$), points ($a$), and horizontal scales ($t$) that correspond to the $t = 0$ and $t = 10$ planet runs, respectively. The symbols are the approximate semi-major axes $a$ computed for each 0 to 2$\phi$ change in azimuth of the planet as explained in the text.}
\label{fig:a}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/planete.eps}
\caption{Plot of the eccentricities $e$, computed for each 0 to 2$\phi$ change in azimuth of the planet as explained in the text, as a function of time. The diamonds and stars (colored red and black in the online version) indicate the points ($e$) and horizontal scales ($t$) that correspond to the $t = 0$ and $t = 10$ planet runs, respectively.}
\label{fig:e}
\end{figure}
